---
title: "RAB117_report"
author: "Marion Hardy"
date: "2023-05-12"
output: 
  html_document:
    toc: true 
    theme: spacelab 
    highlight: monochrome
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, cache = TRUE, echo = FALSE, warning = F, cache.lazy = F)

library(tidyverse)
library(ggrepel)
library(readxl)
library(purrr)
library(msigdbr)
library(clusterProfiler)
library(ReactomePA)
library(org.Mm.eg.db)

```

# Introduction

Analyse d'un screening CRISPR dans des cellules tryptophane neg.
Raphaële a fourni un fichier word détaillé qui explique la manip et les
algorithmes utilisés.

Il y a trois fichiers MLE et trois RRA (triplicats biologiques) qui
contiennent chacun un b score = une moyenne, un z score = un score qui
prend en compte la moyenne et la déviation standard et des pval + wald
tests + fdr.

Je vais pool les trois fichiers, avoir le score z des triplicats et
déterminer les outliers puis les comparer à ceux que Raphaële a (pour
contrôler que je fais pas du caca). Ensuite je prendrai les outliers
recroisés entre le MLE et RRA pour pouvoir produire les différentes
figures dont Raphaële a besoin.

```{r Data loading MLE, include=TRUE, echo=T}

files = list.files("./data/", pattern = "*mle_trp*")

for (i in 1:length(files)){
  file = read_xlsx(paste0("./data/",files[i]))
  file = file %>% filter(sgRNA >=4) %>% 
    dplyr::select(Gene, `treatment|beta`,`treatment|p-value`)
  colnames(file) = c("Gene", paste0("mean_",i), paste0("pval_",i))
  assign(paste0("mle_",i), file)
}

data = list(mle_1, mle_2, mle_3) %>% 
  purrr::reduce(full_join)

data_all = 
  data %>% 
  drop_na() %>% 
  rowwise() %>% 
  mutate(mean = mean(c(mean_1, mean_2, mean_3)),
         sd = sd(c(mean_1, mean_2, mean_3)),
         pval = mean(c(pval_1, pval_2, pval_3))) %>% 
  dplyr::select(Gene, mean, sd, pval)

# Replacing the pvalues = 0 into pval = 10^-5.5 to fit on the graph

data_all[data_all$pval==0,]$pval = 10^-5.5

data = 
  data_all %>% 
    filter(pval<= 0.05)

boxplot(data$mean)
median(data$mean)

outliers = 
  data %>% 
  filter(mean < quantile(data$mean)[2]|mean > quantile(data$mean)[4])

down = 
  data %>% 
  filter(mean < -0.3)

outliers %>% 
  head(15) %>% 
  knitr::kable()

dim(outliers)

```

Quand je filtre pour pval \< 0.05, j'ai 288 outliers pour les MLE
compparé aux 261 de Raphaële. On va partir de la liste des outliers
MLE+RRA qu'elle a fournie pour la suite.

# Possible plots

```{r}

outliers = read_xlsx("./data/RAB117_outliers_MLE+RRA.xlsx", sheet = "Outliers RRA+MLE") %>% 
  dplyr::select(Gene)

outliers = 
  outliers %>% 
  add_row(Gene = c("Gcn1l1", "Slc7a5","Eif2ak4", "Atf4")) # On rajoute les ctrl neg

data1 =
  data %>% 
  filter(Gene%in%outliers$Gene)

```

## MLE : pval and mean

```{r, fig.height=5, fig.width=5}

data_all %>% 
  filter(!is.na(pval)) %>% 
  ggplot(aes(x = mean, y = -log10(pval), label = Gene))+
  geom_point(size = .5, color = 'grey')+
  theme_bw()+
  scale_y_continuous(limits = c(-.2,6)) +
  geom_point(data = data1, size = 1, color = "firebrick") +
  geom_point(data = data1 %>% filter(Gene %in% c("Gcn1l1", "Slc7a5","Eif2ak4","Atf4")), 
             color = "blue", size = 1) +
  geom_text_repel(data = subset(data_all, Gene%in%outliers$Gene),
                  box.padding = 0.25,
                  segment.color = 'grey50',
                  max.overlaps = 30,
                  fontface = "italic")+
  annotate(geom="text", x=-0.75, y=0, label="Negative hits",color="blue")+
  annotate(geom="text", x=0.85, y=0, label="Positive hits",color="firebrick3")+
  annotate(geom="text", x=0.4, y=5, label="p-value < 10e-5",color="black", fontface = "italic")+
   theme(axis.line = element_line(color='black'),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())+
  labs(x = "Mean beta score (n=3)", y = "-Log10(mean(p-value)) (n = 3)",
       title = "MLE+RRA outliers highlighted", subtitle =  "(and negative controls)")

```

```{r, fig.height=3, fig.width=4}

outliers = read_xlsx("./data/RAB117_outliers_MLE+RRA.xlsx", sheet = "Outliers RRA+MLE") %>% 
  dplyr::select(Gene) # Relire le fichier mais sans ajouter les ctrl neg

data1 =
  data %>% 
  filter(Gene%in%outliers$Gene)

data_all %>% 
  filter(!is.na(pval)) %>% 
  ggplot(aes(x = mean, y = -log10(pval), label = Gene))+
  geom_point(size = .5, color = 'grey')+
  theme_bw()+
  scale_y_continuous(limits = c(-.2,6)) +
  geom_point(data = data1, 
             size = .5, color = "firebrick") +
  geom_label_repel(data = subset(data_all, Gene%in%outliers$Gene),
                  box.padding   = 0.35, 
                  point.padding = 0.5,
                  segment.color = 'grey50',
                  max.overlaps = 20)+
  labs(x = "Mean beta score (n=3)", y = "-Log10(mean(p-value)) (n = 3)",
       title = "MLE+RRA outliers highlighted")

```

```{r, fig.height=3, fig.width=4}

negctrl = c("Gcn1l1", "Slc7a5","Eif2ak4")

data1 =
  data %>% 
  filter(Gene%in%negctrl)

data_all %>% 
  filter(!is.na(pval)) %>% 
  ggplot(aes(x = mean, y = -log10(pval), label = Gene))+
  geom_point(size = .5, color = 'grey')+
  theme_bw()+
  scale_y_continuous(limits = c(-.2,6)) +
  geom_point(data = data1, 
             size = .5, color = "blue") +
  geom_label_repel(data = subset(data_all, Gene%in%c("Gcn1l1", "Slc7a5","Eif2ak4")),
                  box.padding   = 0.35, 
                  point.padding = 0.5,
                  segment.color = 'grey50',
                  max.overlaps = 50)+
  labs(x = "Mean beta score (n=3)", y = "-Log10(mean(p-value)) (n = 3)",
       title = "MLE+RRA negative controls highlighted")

```

## RRA : pval and mean

Il faudrait retrouver les fichiers avec les pvaleurs, je peux pas plotter sinon.

## Number of gene attributed at random

```{r}

data_all$number = sample(1:nrow(data_all)) # adding a random number to each gene

data1 =
  data_all %>% 
  filter(Gene%in%outliers$Gene)

data_all %>% 
  ggplot(aes(x = number, y = -log10(pval), label = Gene))+
  geom_point(size = .5, color = 'grey')+
  theme_bw()+
  geom_point(data = data1, 
             size = .5, color = "firebrick") +
  geom_label_repel(data = subset(data_all, Gene%in%outliers$Gene),
                  box.padding   = 0.35, 
                  point.padding = 0.5,
                  segment.color = 'grey50',
                  max.overlaps = 10)
```

## Number of gene attributed not at random

```{r, fig.height=6, fig.width=8}

data_all =
  data_all %>% 
  arrange(pval)

data_all$number = c(1:nrow(data_all)) # addind ordered row numbers

data1 =
  data_all %>% 
  filter(Gene%in%outliers$Gene)

data_all %>% 
  ggplot(aes(x = number, y = -log10(pval), label = Gene))+
  geom_line(size = .5, color = 'black')+
  theme_bw()+
  geom_point(data = data1, 
             size = 1, color = "firebrick") +
  geom_label_repel(data = subset(data_all, Gene%in%outliers$Gene),
                  box.padding   = 0.35, 
                  point.padding = 0.5,
                  segment.color = 'grey50',
                  max.overlaps = 45)
```

# ORA

## Gene ontology

### On the 577 significant genes (not specifically the outliers)

```{r, fig.height=12, fig.width=8}

# Créer la liste de gènes à analyser pour enrichissement

data_sign = data_all %>% filter(pval <= 0.05)

# Set the randomness number (makes it reproducible for anyone)
# Is setting the seed for all the following code to the end of the document

set.seed(054057)

# Check ORA
ora_results_go <- enrichGO(
  gene = data_sign$Gene,
  universe = data_all$Gene,
  keyType = "SYMBOL",
  OrgDb = org.Mm.eg.db,
  ont = "BP",
  pAdjustMethod = "BH")

# Barplot
barplot(ora_results_go, showCategory = 30, order = TRUE,
        x = "Count")+ 
  ggtitle("Enrichment analysis gRNA trp+/-")+
   theme(axis.line = element_line(color='black'),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())

# Table
ora_results_go %>% 
  as_tibble() %>% 
  head(10) %>% 
  dplyr::select(1:6) %>%
  knitr::kable()

```

### On the 177 positive outliers

```{r, fig.height=12, fig.width=8}

data_outliers = data_all %>% filter(pval <= 0.05, Gene%in%outliers$Gene)

ora_results_go1 <- enrichGO(
  gene = data_outliers$Gene,
  universe = data_sign$Gene,
  keyType = "SYMBOL",
  OrgDb = org.Mm.eg.db,
  ont = "ALL",
  pAdjustMethod = "BH")

barplot(ora_results_go1%>% filter(ONTOLOGY == "BP"), showCategory = 30, order = TRUE,
        x = "Count")+ 
  ggtitle("Enrichment analysis gRNA trp+/-")+
   theme(axis.line = element_line(color='black'),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())

dotplot(ora_results_go1 %>% filter(ONTOLOGY == "BP"), showCategory = 30)+ ggtitle("Enrichment analysis gRNA trp+/-")+
   theme(axis.line = element_line(color='black'),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())

ora_results_go1 %>% 
  as_tibble() %>% 
  head(10) %>% 
  dplyr::select(1:6) %>%
  knitr::kable()

```

### On the 144 negative outliers

```{r, fig.height=7, fig.width=7}

data_all %>% 
  filter(!is.na(pval)) %>% 
  ggplot(aes(x = mean, y = -log10(pval), label = Gene))+
  geom_point(size = .5, color = 'grey')+
  theme_bw()+
  scale_y_continuous(limits = c(-.2,6)) +
  geom_point(data = down, size = 1, color = "blue") +
  geom_text_repel(data = subset(data_all, Gene%in%down$Gene),
                  box.padding = 0.25,
                  segment.color = 'grey50',
                  max.overlaps = 30,
                  fontface = "italic")+
  annotate(geom="text", x=-0.75, y=0, label="Negative hits",color="blue")+
   theme(axis.line = element_line(color='black'),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())+
  labs(x = "Mean beta score (n=3)", y = "-Log10(mean(p-value)) (n = 3)",
       title = "MLE+RRA outliers highlighted", subtitle =  "(and negative controls)")
```

```{r, fig.height=12, fig.width=8}

data_outliersdown = data_all %>% filter(pval <= 0.05, Gene%in%down$Gene)

ora_results_go1 <- enrichGO(
  gene = data_outliersdown$Gene,
  universe = data_sign$Gene,
  keyType = "SYMBOL",
  OrgDb = org.Mm.eg.db,
  ont = "ALL",
  pAdjustMethod = "BH")

barplot(ora_results_go1%>% filter(ONTOLOGY == "BP"), showCategory = 30, order = TRUE,
        x = "Count")+ 
  ggtitle("Enrichment analysis gRNA trp+/-")+
   theme(axis.line = element_line(color='black'),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())

dotplot(ora_results_go1 %>% filter(ONTOLOGY == "BP"), showCategory = 30)+ 
  ggtitle("Enrichment analysis gRNA trp+/-")+
   theme(axis.line = element_line(color='black'),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())

ora_results_go1 %>% 
  as_tibble() %>% 
  head(10) %>% 
  dplyr::select(1:6) %>%
  knitr::kable()

```

## Reactome

### On the 577 significant genes (not specifically the outliers)

```{r, fig.height=12, fig.width=8}

data_sign = data_all %>% filter(pval <= 0.05)
annot = read.csv("./data/Biomart_annotations_mm10.csv") # read my annotation file
temp = left_join(data_sign, annot, by = c("Gene" = "gene")) # concatenate the tables

gene_list <-  temp$ENTREZID

ora_results_react <- enrichPathway(gene_list,
                                organism = "mouse",
                                readable = T)

dotplot(ora_results_react)+ 
  ggtitle("Enrichment analysis gRNA trp+/-")+
   theme(axis.line = element_line(color='black'),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())

ora_results_react@result %>% 
  as_tibble() %>% 
  dplyr::select(1:6) %>% 
  head(10) %>% 
  knitr::kable()

```

### On the positive 177 outliers

```{r, fig.height=8, fig.width=8}

temp = left_join(data_outliers, annot, by = c("Gene" = "gene"))
lookup  <-  temp$ENTREZID

## Let's try another way to annotate the genes

library(org.Mm.eg.db)
lookup = select(org.Mm.eg.db, keys = temp$Gene, keytype = "SYMBOL", columns = c("ENTREZID", "SYMBOL"))

ora_results_react1 <- enrichPathway(lookup$ENTREZID,
                                organism = "mouse",
                                readable = T)

dotplot(ora_results_react1)+ 
  ggtitle("Enrichment analysis gRNA trp+/-")+
   theme(axis.line = element_line(color='black'),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())

temp = 
  ora_results_react1@result %>% 
  as_tibble() 

temp %>% 
  dplyr::select(1:6) %>% 
  head(10) %>% 
  knitr::kable()

# library(xlsx)
# write.xlsx(temp,"./data_output/ORA_reactome_177_targets.xlsx")

```

### On the negative 144 outliers

```{r, fig.height=8, fig.width=8}

temp1 = left_join(data_outliersdown, annot, by = c("Gene" = "gene"))
gene_list <-  temp1$ENTREZID

ora_results_react1 <- enrichPathway(gene_list,
                                organism = "mouse",
                                readable = T)

dotplot(ora_results_react1)+ 
  ggtitle("Enrichment analysis gRNA trp+/-")+
   theme(axis.line = element_line(color='black'),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())

ora_results_react1@result %>% 
  as_tibble() %>% 
  dplyr::select(1:6) %>% 
  head(10) %>% 
  knitr::kable()
```

# Highlight genes per found role

## Outliers found in GO terms (BP and CC)

```{r, fig.height=12, fig.width=8}

data_outliers = data_all %>% filter(pval <= 0.05, Gene%in%outliers$Gene)

ora_results_go1 <- enrichGO(
  gene = data_outliers$Gene,
  universe = data_sign$Gene,
  keyType = "SYMBOL",
  OrgDb = org.Mm.eg.db,
  ont = "ALL",
  pAdjustMethod = "BH")

ora_results_go1 %>% 
  as_tibble() %>% 
  head(30) %>% 
  dplyr::select(1:6) %>%
  knitr::kable()

```

Organize by theme

-   BP Mitochondrial gene expression (encompasses mitochondrial
    translation)

-   BP ATP metabolic process (encompasses proton-motive force driven ATP
    prod)

-   CC respiratory chain complex I

-   BP oxidative phosphorylation

## Outliers found in reactome

```{r, fig.height=8, fig.width=8}

temp = left_join(data_outliers, annot, by = c("Gene" = "gene"))
gene_list <-  temp$ENTREZID

ora_results_react1 <- enrichPathway(gene_list,
                                organism = "mouse",
                                readable = T)

ora_results_react1@result %>% 
  as_tibble() %>% 
  dplyr::select(1:4) %>% 
  head(30) %>% 
  knitr::kable()

```

## Organize and highlight targets by theme and functions

```{r}

outliers = 
  outliers %>% 
  add_row(Gene = c("Gcn1l1", "Slc7a5","Eif2ak4", "Atf4"))

data1 =
  data %>% 
  filter(Gene%in%outliers$Gene)

# Get the genes listed in position x of reactome ORA results
translation = unique(unlist(c(strsplit(ora_results_react1$geneID[[1]],"/"),
  strsplit(ora_results_react1$geneID[[2]],"/"),
  strsplit(ora_results_react1$geneID[[3]],"/"),
  strsplit(ora_results_react1$geneID[[4]],"/"))))

TCA = unique(unlist(c(strsplit(ora_results_react1$geneID[[5]],"/"),
  strsplit(ora_results_react1$geneID[[6]],"/"),
  strsplit(ora_results_react1$geneID[[7]],"/"),
  strsplit(ora_results_react1$geneID[[8]],"/"))))

other = unique(unlist(c(strsplit(ora_results_react1$geneID[[9]],"/"),
  strsplit(ora_results_react1$geneID[[10]],"/"),
  strsplit(ora_results_react1$geneID[[11]],"/"),
  strsplit(ora_results_react1$geneID[[12]],"/"),
  strsplit(ora_results_react1$geneID[[13]],"/"))))

the_rest = outliers$Gene[!(outliers$Gene %in% c(translation, TCA, other))]

labels = c("Mrpl28","Ndufa2","Ndufc2","Mrps25","Atp5b","Ndufa8",
           "Cand1","Gcn1l1", "Slc7a5","Eif2ak4","Atf4","Mrpl45", "Atp5o",
           "Mrpl20","Ndufa1","Gadd45gip1","Cars2","Ears2")
  
```


```{r}

data_all %>% 
  filter(!is.na(pval)) %>% 
  ggplot(aes(x = mean, y = -log10(pval), label = Gene))+
  geom_point(size = .5, color = 'grey90', alpha = 0.4)+
  theme_bw()+
  scale_y_continuous(limits = c(-.2,6)) +
  geom_point(data = data1 %>% filter(Gene %in% the_rest), 
             color = "grey70", size = 2) +
  geom_point(data = data1 %>% filter(Gene %in% translation), 
             fill = "#ffa73e", size = 2, shape = 21) +
  geom_point(data = data1 %>% filter(Gene %in% TCA), 
             fill = "lightskyblue1", size = 2, shape = 21, alpha = 1) +
  geom_point(data = data1 %>% filter(Gene %in% other), 
             fill = "grey50", size = 2, shape = 21, fill = "white") +
  geom_point(data = data1 %>% filter(Gene %in% c("Gcn1l1", "Slc7a5","Eif2ak4","Atf4")), 
             color = "grey70", size = 2) +
  geom_text_repel(data = subset(data_all, Gene%in%labels),
                  box.padding = 0.4,
                  segment.color = 'grey50',
                  segment.size = 0.2,
                  max.overlaps = 10,
                  fontface = "italic",
                  size = 3,
                  point.size = 10,
                  min.segment.length = 0)+
  annotate(geom="text", x=-0.75, y=0, label="Negative hits",color="black")+
  annotate(geom="text", x=0.85, y=0, label="Positive hits",color="black")+
  annotate(geom="text", x=-0.8, y=5.5, label="p-value < 10e-5",color="black", 
           fontface = "italic", size = 3)+
   theme(axis.line = element_line(color='black'),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())+
  labs(x = "Mean beta score (n=3)", y = "-Log10(mean(p-value)) (n = 3)",
       title = "MLE+RRA outliers highlighted", subtitle =  "(and negative controls)")

ggsave(plot = last_plot(),"./figures/MLE_RRA_ORA_targets.svg", device = "svg",
       dpi = 300, height = 4, width = 4)

ggsave(plot = last_plot(),"./figures/MLE_RRA_ORA_targets.png", device = "png",
       dpi = 300, height = 4, width = 5.5)
```

```{r}
sessionInfo()
```
